A Review of Mortality Models

### 2020-09-07 mortality-models/draft.tex
WIP 07Sep20

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.
\bibliography{draft}
\bibliographystyle{plain}
\end{document}

>>>>

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.

\section{Heligman-Pollard model}\label{s5}
The term `odds' is a well-defined term in probability theory. When we are 
dealing with an experiment with two outcomes - that something happens or 
it does not - then it is common in certain fields to express the 
probabilities in terms of odds.  For example, if $p$ is the probability 
that an event will happen and $q = 1 - p$ is the probability that it will 
not happen then the odds of the experiment is $p/q$.

Recall from equations \eqref{s1e2} and \eqref{s1e3} that $p_x$ is the 
probability that life aged $x$ will survive for another year and $q_x$ is
the probability of the complimentary event. Therefore, the odds of death in 
this case are $q_x/p_x$.  Heligman and Pollard \cite{heligman1980age} proposed 
a model that has the algebraic structure of Thiele's mortality law but 
applied to odds of survival. Mathematically it is expressed as
\begin{equation}\label{s5e1}
\frac{q_x}{p_x} = A^{(x+B)^C}+
D\exp\left(-E\log^2\left(\frac{x}{F}\right)\right) + GH^x.
\end{equation}
The model has eight parameters, $A, B, C, D, E, F, G$ and $H$. Of these, $A,
B, C, D \in (0, 1)$, $E > 0$, $F \in (15, \omega)$, $G \in (0, 1)$ and $H
> 0$. The constant $\omega$ is the hump maximum and it can be between $55$
and $100$. The Heligman-Pollard model was first developed for the Australian
population post the second world war. An advantage of Heligman-Pollard model 
is in biological interpretation of each of the three terms on the right hand 
side of equation \eqref{s5e1}.

The first term on the right hand side of equation \eqref{s5e1} represent 
infantile mortality. It is a fast declining exponential representing the 
rapid fall in mortality during infancy. $A$ is approximately equal to $q_1$ 
and is a measure of the mortality rate of a one year old child. $C$ is a 
measure of how quickly the new-born adjusts to its surroundings and gains 
immunity. A higher $C$ indicates the fast decline in infant mortality as the
child passes through infancy. When $B = 0$, $q_0 = 0.5$ irrespective of the
values taken by $A$ and $C$. The value of $q_0$ is closer to $q_1$ if $B$ is
higher for a given value of $C$. $B$ is thus an indicator of the central 
tendancy of infant mortality.

The second term is gaussian in the logarithm of age. It represent the accident
mortality for young adults of both sexes and maternal mortality in young 
females. It covers the `accident hump` observed in all demographic data. It
is prominently observed between ages $10$ and $40$. The parameter $D$ controls 
the severity of accident hump, $E$ determines the spread of the hump and $F$ 
the location.  Using a statistical analogy, $E$ is a measure of spread and 
$F$ that of the central tendency.

The third term traces its roots to the Gompertz model. It represents the
exponential increase in the deterioration of the body. The parameter $G$ 
represents the senescent mortality and $H$ the increase in the rate of that
process.

\bibliography{draft}
\bibliographystyle{plain}
\end{document}
